How to select predictive models for decision-making or causal inference?

### 2024-06-02 how_to_select.tex
format figure for new format

% added commands and packages
\newtheorem{assumption}{Assumption}
\newtheorem{proposition*}{Proposition}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\eg}{\emph{e.g.}}
\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
\usepackage{mathtools}
\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
\usepackage{booktabs} % For formal tables (eg. bottomrules command)
\usepackage{threeparttable} % For table 
\usepackage{caption} % for subfigures
\usepackage{subcaption}
\usepackage{placeins} % for FloatBarrier command to prevent a figure to pass some point.

@@

\abstract{\textbf{Background:}
    We investigate which procedure selects the predictive model most
    trustworthy to reason on the effect of an intervention and support
    decision making.\\
    %As predictive models --\eg from machine learning-- give likely
    %outcomes, they may be used to, a
    %causal-inference task. The increasing complexity of health data has opened
    %the door to a plethora of models, but also the Pandora box of model
    %selection: which of these models yield the most valid causal estimates? 
    %\textbf{Methods:}
    We study such a variety of model selection procedures in practical settings:
    finite samples settings and without theoretical assumption of
    \emph{well-specified} models. Beyond standard cross-validation or
    internal validation procedures, we also study elaborate causal risks.
    These build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. We evaluate whether
    empirically estimated nuisances, which are necessarily noisy,
    add noise to model selection. We compare different metrics for
    causal model selection in an extensive empirical
    study based on a simulation and three healthcare datasets
    based on real covariates. \\
    \textbf{Results:} Among all metrics, the mean squared error,
    classically used to evaluate predictive modes,
    is worse. Re-weighting it with propensity score
    does not bring much improvements. The
    $R\text{-risk}$, which uses as nuisances a model of
    mean outcome and propensity scores, leads to the best performances.
    Nuisance corrections are best estimated with flexible estimators such
    as a super learner.
    \\
    \textbf{Conclusions:} When predictive models are used to
    reason on the effect of an intervention, they must be evaluated
    with different procedures than standard predictive
    settings; using the $R\text{-risk}$ from causal inference.}
\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }

@@

\begin{figure}[h!]
    % XXX: for submission
    \centering\begin{minipage}{.6\linewidth}
        {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
                causal inference}}

@@

        {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
            prediction but better causal inference}}

@@

    \centering\begin{minipage}{.6\linewidth}
        \includegraphics[width=\linewidth]{estimation_procedure_causal_selection_procedure.png}
    \end{minipage}
    \caption{Estimation procedure for causal model
        selection.}\label{problem:estimation_procedure:figure}
\end{figure}

@@

\begin{figure}[!b]
    % XXX: for submission
    \centering\begin{minipage}{.75\linewidth}
        \includegraphics[width=\linewidth]{_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
    \end{minipage}
    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
        \ref{apd:table:relative_kendalls_all_datasets} gives median and
        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
\end{figure}

@@

\begin{figure}[!h]
    % XXX: for submission
    \centering\begin{minipage}{.75\linewidth}
        \includegraphics[width=\linewidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
    \end{minipage}
    \caption{\textbf{Model selection is harder for low population
            overlap}:
        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
        Kendall's and continuous overlap values in Figure
            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
\end{figure}

@@

\begin{figure}[!tb]
    % XXX: for submission
    \centering\begin{minipage}{.75\linewidth}
        \includegraphics[width=\linewidth]{_3_procedure_r_risk_only_3datasets_twocols.pdf}
    \end{minipage}

@@

\begin{figure}[!tb]
    % XXX: for submission
    \centering\begin{minipage}{.75\linewidth}
        \includegraphics[width=\linewidth]{_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
    \end{minipage}

@@

    \begin{figure}[!t]
        \centering
        \includegraphics[width=\linewidth]{caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}

>>>>

% added commands and packages
\newtheorem{assumption}{Assumption}
\newtheorem{proposition*}{Proposition}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\eg}{\emph{e.g.}}
\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
\usepackage{mathtools}
\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
\usepackage{booktabs} % For formal tables (eg. bottomrules command)
\usepackage{threeparttable} % For table 
\usepackage{caption} % for subfigures
\usepackage{subcaption}
\usepackage{placeins} % for FloatBarrier command to prevent a figure to pass some point.
\usepackage{titlesec} % for title formatting
\titleformat*{\subsection}{\color{jnlclr}\bfseries} % force colored subsection font for paragraphs
\titleformat*{\paragraph}{\bfseries} % force bold font for paragraphs

@@

\abstract{\textbf{Background:}
    We investigate which procedure selects the predictive model most
    trustworthy to reason on the effect of an intervention and support
    decision making.\\
    %As predictive models --\eg from machine learning-- give likely
    %outcomes, they may be used to, a
    %causal-inference task. The increasing complexity of health data has opened
    %the door to a plethora of models, but also the Pandora box of model
    %selection: which of these models yield the most valid causal estimates? 
    \textbf{Methods:}
    We study such a variety of model selection procedures in practical settings:
    finite samples settings and without theoretical assumption of
    \emph{well-specified} models. Beyond standard cross-validation or
    internal validation procedures, we also study elaborate causal risks.
    These build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. We evaluate whether
    empirically estimated nuisances, which are necessarily noisy,
    add noise to model selection. We compare different metrics for
    causal model selection in an extensive empirical
    study based on a simulation and three healthcare datasets
    based on real covariates. \\
    \textbf{Results:} Among all metrics, the mean squared error,
    classically used to evaluate predictive modes,
    is worse. Re-weighting it with propensity score
    does not bring much improvements. The
    $R\text{-risk}$, which uses as nuisances a model of
    mean outcome and propensity scores, leads to the best performances.
    Nuisance corrections are best estimated with flexible estimators such
    as a super learner.
    \\
    \textbf{Conclusions:} When predictive models are used to
    reason on the effect of an intervention, they must be evaluated
    with different procedures than standard predictive
    settings; using the $R\text{-risk}$ from causal inference.}
\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }

@@

\begin{figure}[h!]
    % XXX: for submission
    \centering\begin{minipage}{.95\linewidth}
        {\sffamily\footnotesize\scalebox{0.95}{a) Random forest, good average prediction but bad
                causal inference}}

@@

        {\sffamily\footnotesize\scalebox{0.95}{b) Linear model, worse average
            prediction but better causal inference}}

@@

    \centering\begin{minipage}{.95\linewidth}
        \includegraphics[width=\linewidth]{estimation_procedure_causal_selection_procedure.png}
    \end{minipage}
    \caption{Estimation procedure for causal model
        selection.}\label{problem:estimation_procedure:figure}
\end{figure}

@@

\begin{figure}[!b]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
    \end{minipage}
    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
        \ref{apd:table:relative_kendalls_all_datasets} gives median and
        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
\end{figure}

@@

\begin{figure}[!h]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
    \end{minipage}
    \caption{\textbf{Model selection is harder for low population
            overlap}:
        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
        Kendall's and continuous overlap values in Figure
            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
\end{figure}

@@

\begin{figure}[!tb]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_3_procedure_r_risk_only_3datasets_twocols.pdf}
    \end{minipage}

@@

\begin{figure}[!tb]
    % XXX: for submission
    \centering\begin{minipage}{\linewidth}
        \includegraphics[width=\linewidth]{_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
    \end{minipage}

@@

    \begin{figure}[!t]
        \centering
        \includegraphics[width=0.8\linewidth]{caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
